A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-01-04 analysis/paper/paper.Rmd
editing through to conlcusion

---
title: "A Bayesian approach to burial networks to infer social changes in 17th northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationship between entities. Different from traditional static networks, statistical approaches to network analysis, such as exponential random graph models (ERGMs), provides a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduced a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a more centralized structure with an emphasis of social differentiation after the European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expand the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  These are the highlights. 
---

@@

@@

<!-- archaeological specific case study -->
In this study, we explore the changes in social inequality in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that the social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the network of our hypothesized model? (2) did European colonial activities in 17th century Taiwan result in an increased social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of statistical modeling for burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

@@

<!--Model fitting--> 
After we set our model parameters, we simulated networks in a Bayesian framework using a Markov chain Monte Carlo (MCMC) algorithm. MCMC algorithms allow estimation of posterior distributions through direct random sampling the posterior without assuming the prior comes from any specific distribution [@Hamra2013]. We can obtain a posterior distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on an algorithm. This enables a random or stochastic simulation in a long run where each move does not depend on the previous move. More chains ensures a more desirable posterior distribution that is close to the target distribution under study, or convergence. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and then switches to a pair of actors at random within the chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations was `r burn_in` and the number of iterations after the burn-in was `r main_iter`. We set the number of iterations used to simulate a network y′ at each iteration to `r aux_iter`.

@@

<!-- theoretical framework for the results-->
A striking finding in our results is the change in network properties of Kiwulan burials from a more cohesive network with multiple subgroups (represented by higher transitivity and negative popularity) toward a more centralized network with concentration in a limited burials (lower transitivity and positive popularity) after the European presence during the 17th century. This supports our hypothesized models of chronological change from a more corporate society to a more networked society, as indicated in the grave goods and burial attributes. These observations are in line with theoretical assumptions of horizontal hierarchy, the corporate-network continuum that focuses on distinct strategies for achieving power as an alternative explanation to understand the emergence and level of social inequality [@Drennan2010; @Feinman2000; @Feinman2000political]. Compared to traditional models of vertical hierarchies that stress stratified social characteristics, the corporate-network model may be more relevant for understanding how a small scale society was organized, such as we see here in the pericolonial context of Kiwulan. A corporate-based society, such as the pre-European situation at Kiwulan, stresses shared power between individuals, communal  rituals, and social inequality, if any, would be associated with a group association [@Siegel1999]. In contrast, a more network-based society, indicated by the post-European data from Kiwulan, presents wealth accumulation through individual networks, prestige good system manipulation, and trade monopolization [@Blanton1996]. It should be noted that corporate/network continuum represents a dynamic process with different degree of hierarchical complexity instead of a static ideal-type stage.

<!--archaeological implication based on the results-->
The covariants for both networks, before and after the European presence, provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:xxx)). We included age, sex, ritual pottery, and a wealth index based on the value of grave goods. The pre-European network shows that burials with pottery as ritual vessels tend to be tied. On the other hand, in the post-European network ties tend to be between burials within the same level of wealth. Moreover, in the post-European network, burials with ritual vessels tend not to be tied. This illustrates that different social and economic mechanisms determined relations between burials for the two periods. Before European arrival, ritual behaviors seem to have been important status indicators, relative to the use of prestige goods. After European presence was established, status indicators shifted from ritual to wealth differences. Distance between burials did not correlate to network formation for both periods that means geographical proximity, implying some kinship association, is not a factor for sharing similar goods between burials. This may suggest that the use of burial goods is more individual-oriented instead of kinship-based sharing. In addition, demographic information, such as age and sex, do not show any correlations with network formation for either period. This may be due to the unavailability of age and sex values for many burials in our sample.

@@

<!-- other evidence -->
We used burial data as a proxy to explore the social relations based on the argument that burials could be viewed as a social representation to understand social structures [@Binford1971; @Saxe1970]. We note that there could be an issue of disconnection between the role of the deceased in a burial context and their real life, which should be reduced by examining evidence from the residential context [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution when the Europeans were active in northern Taiwan that hints at an increased social differentiation at Kiwulan in a pericolonial context [@LWandBM2020ornament]. This is consistent with the result of a more centralized burial network after the presence of the Europeans, which supports the assumption of social role representation from burials in this case. Since the burial goods used for making ties were treated as prestige goods or status items in local Indigenous culture [@Borao2009; @LiandWu2006], the interrelations represented by the flow of goods could reflect the process of social changes at Kiwulan in a pericolonial context.

<!--fit in a bigger historical picture in a periclonial context -->
The involvement in long-distance exchange is often associated with changes in political-economic strategies to a network mode, where ambitious individuals are able to build their local power base and personal influence through the distribution of high-value goods [@Blanton1996; @Feinman2000; @Klehm2017]. Northeastern Taiwan was involved in a regional exchange in East Asia since the 14th century and long-distance trade during the European presence in the 17th century [@Chen2005; @Wang2007]. In 1626, the Spanish built their forts in Keelung and Tamsui, northern Taiwan, and later defeated and took over by the Dutch in 1642 [@Andrade2007]. Indigenous people in northeastern Taiwan first encountered colonial power in 1632 when they were attached by the Spanish and later in 1647 by the Dutch, who also asked them to pay annual tributes in a flexible way [@Borao2009; @Kang2016]. The Dutch would give local Indigenous leaders symbolic items or trade goods to assert their political authority and consolidate their relationship with Indigenous societies.

<!--anthopological implication at Kiwulan and limitation-->
At Kiwulan, after the European presence, some individuals may be differentiated or privileged to possess more distinctive goods through becoming the trade partners with the Europeans. Changes in interburial networks we observed may result from differential access to exotic trade relative to the European colonization and gradually led to an increased social differentiation. The image of European power could be embedded into prestige objects pre-existing in Indigenous culture where their values were amplified with concepts of wealth or power, since good values are contextually mutable and entangled in historical processes [@Aswani2003; @Thomas2009]. This may lead to ambitious individuals to compete with each other for the goods and accumulate wealth. It also demonstrates the agency of local Indigenous societies to incorporate and manipulate trade items in a precolonial context with weak colonial control, gradually resulting in social changes from a more corporate to a more network strategy. Our interpretation is based on relations indicated by burial data that can be supported by the clustered distribution of trade goods in Kiwulan residential area [eg. @LWandBM2020ornament]. We would still need more other different evidence to understand the process of social change. In addition, we acknowledge the fragmentation issue in archaeology data that might limit network interpretation [eg. @Erik2015], such as missing age and sex nodal attributes for some burials in our data. Despite we lack of informative results about their effect on network formation, the Bayesian approach still provides effective and reliable statistical inferences on the overall network properties by reducing the missing data effect.

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
In this paper, we presented a naval approach for studying burial relations using ERGM network analysis within a Bayesian framework to rigorously examined social changes in a European pericolonial context in northeastern Taiwan. We tested the hypotheses of chronological changes in interburial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our hypothesize model that the relationship between burials changed after European colonization period in the 17th century. Before the arrival of the Europeans, the network properties has a tendency of more closed burial relationships with ritual elements as the key formation mechanism. After the European arrival, the network has a tendency of centralization relative to the rarity of goods. The changes in formation mechanisms for networks from ritual-oriented practices to wealth-oriented possession suggest the behavior of wealth accumulation and an increased differentiation stimulated by the European presence and associated long-distance trade network. This aliens with changes in social complexity from a more corporate to a more network mode society that stress different strategies for achieving power and different degrees of social inequality, indicative of an increased social inequality. Using burial data with historical documents, we are able to detect the processes of changes in Indigenous societies in the periphery of European colonial bases to better understand indirect impacts that are underestimated in this region [@Acabado2017; @Trabert2017]. This study not only discuss pericolonal impacts that are not well understood yet, but also points out the agency of local people to manipulate colonial power instead of being passive recipients of trade goods.

<!-- technical and methodological implication-->
This case study highlights the methodological improvement that Bayesian inference on ERGM can inform and enhance studies in rational data in archaeology. A Bayesian framework can reduce the effects of small sample size or missing data commonly present in archaeological data by incorporating prior information and MCMC estimations. Moreover, probabilistic inference provides robust results of interrelations that are not statistically tested by traditional network approaches. Bayesian network modeling can be applied to a wide range of archaeological data to examine the formation of relationships. This allows in-depth discussion of dynamic processes of relationship formation and their underlying factors that are informative of historical trajectories or socio-cultural phenomena.

>>>>

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationship between entities. Different from traditional static networks, statistical approaches to network analysis, such as exponential random graph models (ERGMs), provides a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  These are the highlights. 
---

@@

# A Case Study from Northereastern Taiwan

@@

<!--fit in a bigger historical picture in a periclonial context -->
The involvement in long-distance exchange is often associated with changes in political-economic strategies to a network mode, where ambitious individuals are able to build their local power base and personal influence through the distribution of high-value goods [@Blanton1996; @Feinman2000; @Klehm2017]. Northeastern Taiwan was involved in a regional exchange in East Asia since the 14th century and long-distance trade during the European presence in the 17th century [@Chen2005; @Wang2007]. In 1626, the Spanish built their forts in Keelung and Tamsui, northern Taiwan, and later were defeated and taken over by the Dutch in 1642 [@Andrade2007]. Indigenous people in northeastern Taiwan first encountered colonial power in 1632 when they were attacked by the Spanish and later in 1647 by the Dutch, who also asked them to pay annual tributes in a flexible way [@Borao2009; @Kang2016]. The Dutch would give local Indigenous leaders symbolic items or trade goods to assert their political authority and consolidate their relationship with Indigenous societies.

<!-- archaeological specific case study -->
In this study, we explore the changes in social inequality in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that the social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the network of our model? (2) did European colonial activities in 17th century Taiwan result in an increased social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of statistical modeling for burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

@@

<!--Model fitting--> 
After we set our model parameters, we simulated networks in a Bayesian framework using a Markov chain Monte Carlo (MCMC) algorithm. MCMC algorithms allow estimation of posterior distributions through direct random sampling the posterior without assuming the prior comes from any specific distribution [@Hamra2013]. We can obtain a posterior distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on an algorithm. This enables a random or stochastic simulation in a long run where each move does not depend on the previous move. More chains ensures a more desirable posterior distribution that is close to the target distribution under study, or convergence. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and then switches to a pair of actors at random within the chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations was `r burn_in` and the number of iterations after the burn-in was `r main_iter`. We set the number of iterations used to simulate a network y' at each iteration to `r aux_iter`.

@@

<!-- theoretical framework for the results-->
A striking finding in our results is the change in network properties of Kiwulan burials from a more cohesive network with multiple subgroups (represented by higher transitivity and negative popularity) toward a more centralized network with concentration of connections among fewer burials (lower transitivity and positive popularity) after the European presence during the 17th century. This supports our models of chronological change from a more corporate society to a more networked society, as indicated in the grave goods and burial attributes. These observations are in line with theoretical assumptions of horizontal hierarchy, the corporate-network continuum that focuses on distinct strategies for achieving power as an alternative explanation to understand the emergence and level of social inequality [@Drennan2010; @Feinman2000; @Feinman2000political]. Compared to traditional models of vertical hierarchies that stress stratified social characteristics, the corporate-network model may be more relevant for understanding how a small scale society was organized, such as we see here in the pericolonial context of Kiwulan. A corporate-based society, such as the pre-European situation at Kiwulan, stresses shared power between individuals, communalrituals, and social inequality, if any, would be associated with groups [@Siegel1999]. In contrast, a more network-based society, indicated by the post-European data from Kiwulan, presents wealth accumulation through individual networks, prestige good manipulation, and trade monopolization [@Blanton1996]. It should be noted that the corporate/network continuum represents a dynamic process with different degrees of hierarchical complexity instead of static ideal-type stages.

<!--archaeological implication based on the results-->
The covariants for both networks, before and after the European presence, provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:xxx)). Our covariants were age, sex, ritual pottery, and a wealth index based on the value of grave goods. The pre-European network shows that burials with pottery as ritual vessels tend to be tied. On the other hand, in the post-European network ties tend to be between burials within the same level of wealth. Moreover, in the post-European network, burials with ritual vessels tend not to be tied. This illustrates that different social and economic mechanisms determined relations between burials for the two periods. Before European arrival, ritual behaviors seem to have been important status indicators, relative to the use of prestige goods. After European presence was established, status indicators shifted from ritual to wealth differences. Physical distance between burials did not correlate to network formation for either period. This means geographical proximity, implying some kinship association, is not a factor for sharing similar goods between burials. This may suggest that the use of burial goods is more individual-oriented instead of kinship-based. In addition, demographic information, such as age and sex, do not show any correlations with network formation for either period. This may be due to the unavailability of age and sex values for many burials in our sample.

<!-- other evidence -->
We used burial data as a proxy to explore the relations based on the argument that burials  represent social structures [@Binford1971; @Saxe1970]. We note that there could be an issue of disconnection between the role of the deceased in a burial context and their life, which could be reduced by examining evidence from the residential context [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution when the Europeans were active in northern Taiwan that hints at an increased social differentiation at Kiwulan in a pericolonial context [@LWandBM2020ornament]. This is consistent with the result of a more centralized burial network after the presence of Europeans, which supports a connection between a burial and the social role of the living person. Since the burial goods used for making ties were treated as prestige goods or status items in the local Indigenous culture [@Borao2009; @LiandWu2006], the interrelations represented by the flow of goods observed in the burial data likely reflect social contexts of the living people at Kiwulan in a pericolonial context.

<!--anthopological implication at Kiwulan and limitation-->
At Kiwulan, changes in interburial networks may result from differential access to exotic trade enabled by European contact, gradually leading to increased social differentiation. The image of European power could be embedded into prestige objects pre-existing in local Indigenous culture where their values were amplified with concepts of wealth or power, since their values are contextually mutable and entangled in historical processes [@Aswani2003; @Thomas2009]. This may lead to ambitious individuals to competing with each other for traditional goods and accumulating wealth. It also demonstrates the agency of local Indigenous societies to incorporate and manipulate trade items in a pericolonial context with weak colonial control, gradually resulting in social changes from a more corporate to a more network strategy. We recognise that the meaning of the burial goods remains speculative without rich ethnographic data to support our interpretations. Furthermore, a detailed material culture context for burial goods in residential settings would also help to strengthen our interpretations, but this is not available from Kiwulan.  In addition, we acknowledge the fragmentation issue in our archaeological data that limit network interpretations [eg. @Erik2015], such as missing age and sex attributes for some burials at Kiwulan.

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
In this paper, we presented a novel approach for studying burial relations using ERGM network analysis within a Bayesian framework. We examined social changes in a European pericolonial context in northeastern Taiwan. We tested the hypotheses of chronological changes in interburial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our model that the relationship between burials changed after the European colonization period in the 17th century. Before the arrival of Europeans, the burial network has a tendency of more closed relationships with ritual elements as the key formation mechanism. After European arrival, the network has a tendency of centralization relative to the rarity of goods. The changes in formation mechanisms for networks from ritual-oriented practices to wealth possession suggest increased behavior of wealth accumulation and differentiation stimulated by the European presence and associated long-distance trade network. This aligns with changes in social complexity from a more corporate to a more network mode society with different strategies for achieving power and different degrees of social inequality. Using burial data with historical documents, we are able to detect changes in Indigenous societies indicative of increased social inequality to better understand indirect European colonial impacts that have been underestimated in this region [e.g. @Acabado2017; @Trabert2017]. Furthermore, our study also highlights the agency of local people to manipulate colonial power instead of being passive recipients of trade goods.

<!-- technical and methodological implication-->
This case study demonstrates the methodological improvement that Bayesian inference on ERGM provides to inform and enhance studies of relational data in archaeology. A Bayesian framework can reduce the effects of small sample size or missing data commonly present in archaeological data by incorporating prior information and MCMC estimations. Bayesian network modeling can be applied to a wide range of archaeological data to examine the formation of relationships using robust probabilistic inference. This enables insights into the dynamic processes of relationship formation and the underlying factors of historical trajectories of socio-cultural phenomena.

@@
